Circumnuclear Media and Accetion Rates of Quiescent Supermassive Black Holes

### 2015-02-10 ms.tex
more edits 3.3

Radiative cooling contributes an additional term to equation (\ref{eq:dsdt}) of the form
\be
\dot{q}_{\rm rad} = -\Lambda(T)n^{2},
\label{eq:qdot_rad}
\ee
where $n = \rho/\mu m_p$ is the gas number density and $\Lambda(T)$ is the cooling function.  We neglect radiative cooling in our baseline calculations, despite the fact that this is not justified when the wind heating $\vwO$ is low or if the mass return rate $\eta$ is high.  However, once radiative cooling becomes important compared to other sources of heating and cooling, its presence leads to thermal instability (e.g.~\citealt{McCourt+12}) that cannot be accounted for by our 1D
time-independent model.  A main goal of this work is to use analytic and numerical solutions neglecting radiation to ascertain over what range of physical conditions cooling instabilities are most likely to develop ($\S\ref{sec:cooling}$).

@@

In the opposite limit of weak heating ($v_{w} \ll \sigma$), the stagnation radius moves to large radii $\gg r_{\rm soi}$ or may not exist at all (formally implying that the entire ISM of the galaxy is accreting).  This result makes intuitive sense: gas is supplied to the nuclear by stars which are gravitationally bound to the black hole, so outflows are possible only if the specific heating rate $\sim \tilde{v}_{w}^{2}$ exceeds the specific gravitational binding energy $\sim \sigma^{2}$.

@@

The accretion rate onto the SMBH is given by the integrated mass loss rate interior to the stagnation radius (eq.~[\ref{eq:q}]), 
\begin{eqnarray}
  \dot{M} &=& \frac{\eta \Menc|_{r_{\rm s}}}{\th} \nonumber \\
&\approx&
  \begin{cases}
    2.9 \times 10^{-5} M_{\bullet,8}^{1.76}
    v_{500}^{-3.8}  \eta_{0.02} \Msun \, \pyear& \text{core} \\
    5.3 \times 10^{-5} M_{\bullet,8}^{1.48} 
    v_{500}^{-2.4}  \eta_{0.02} \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}
where we have made use of equation (\ref{eq:stag_simple}), i.e. assuming that $v_{w} \gg \sigma$, and we have divided our results into core ($\Gamma = 0.1$) and cusp ($\Gamma = 0.8$) galaxies.  The resulting Eddington ratio $\dot{M}/\dot{M}_{\rm edd}$ is given by 
\begin{eqnarray}
\frac{\dot{M}}{\dot{M}_{\rm edd}} &\approx&
  \begin{cases}
    1.3 \times 10^{-5} M_{\bullet,8}^{0.76}
    v_{500}^{-3.8}  \eta_{0.02}   & \text{core}, \\
    2.4 \times 10^{-5} \Mbheight^{0.48} 
    v_{500}^{-2.4}  \eta_{0.02}   & \text{cusp}, 
  \end{cases}
  \label{eq:eddr_analytic}
\end{eqnarray}
where $\dot{M}_{\rm edd} = 2.6M_{\bullet,8}M_{\odot}$ yr$^{-1}$  is the Eddington accretion rate, assuming a radiative efficiency of ten per cent.  Note the sensitive inverse dependence of the accretion rate on the wind heating rate.  
%To reproduce the estimated values $\dot{M}/\dot{M}_{\rm edd} \lesssim 10^{-3}$ in low-luminosity AGN or quiescent galaxies, we thus require $v_{w} \gtrsim 150M_{\bullet,8}^{0.5}$

@@

Our expression for the mass accretion rate (eq.~[\ref{eq:mdot_analytic}]) can be compared to the traditional Bondi rate for accretion onto a point source from an external medium of specified density and temperature (\citealt{Bondi52}):
\begin{align}
  \dot{M}_{\rm B} =4\pi \lambda r_{\rm B}^2 \rho|_{r_{\rm B}}v_{\rm ff}|_{r_{\rm B}},
\label{eq:bondi}
\end{align}
where $r_{\rm B} \equiv GM/c_{\rm s,ad}^{2}$ is the Bondi radius, $c_{\rm s,ad} = \sqrt{\gamma kT/\mu m_p}$ is the adiabatic sound speed, $v_{\rm ff}|_{r_{\rm B}} = r_{\rm B}/t_{\rm ff}|_{r_{\rm B}} = (GM_{\bullet}/r_{\rm B})^{1/2}$ and $\lambda$ is a parameter of order unity.  One observes a similarity between equation (\ref{eq:mdot_gas}) and the Bondi formula (eq.~[\ref{eq:bondi}]) if $R_{\rm B}$ is replaced by $\rs$.  When the stagnation radius resides interior to the sphere of influence we indeed have (eq.~[\ref{eq:stag_simple}])
\begin{align}
  \rs\approx\frac{7}{2}\frac{G \Mbh}{n \vwO^2} \approx \frac{7}{6n}r_{\rm B},
  \label{eq:rs_simple}
\end{align}
where the second equality makes use of equation (\ref{eq:Tanalytic}).  {\bf BDM: please check, as I found something different than what was here before}.

@@

Our analytic results neglect the effects of heat conduction, an assumption we now check.  The ratio of the magnitude of the conductive heating rate to the nominal external (``wind") heating rate at the stagnation radius is given by
\begin{align}
  \left.\frac{\nabla\cdot(\kappa \nabla T)}{q v_{\rm
w}^{2}/2}\right|_{r_{\rm s}} &\sim \frac{2t_{\rm h}\kappa_0
T|_{r_{\rm s}}^{7/2}}{r_{\rm s}^{2}\eta \rho_{\star}|_{r_{\rm s}} \tilde{v}_{w}^{2}(1+\phi)}
\nonumber \\ &\sim {\rm min}
  \begin{cases}
  10 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.8} v_{500}^{6.8} &  \text{unsaturated (core)}\\
 10 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.5} v_{500}^{5.4} &  \text{unsaturated (cusp)}\\
  1.0(\phi/0.1)/(2-\Gamma) & \text{saturated},
  \end{cases}
 \label{eq:rhors}
\end{align}
where the second equality makes use of equations (\ref{eq:Tanalytic}), (\ref{eq:stag_simple}), and we have approximated $\nabla^{2} \sim 1/r_{\rm s}^{2}$, $\nabla \sim 1/r_{\rm s}$.  Because the stagnation radius is usually well inside the break radius, we have approximated the stellar density as (eq.~[\ref{eq:rhostar}])
\begin{eqnarray}
  \rho_{\star}|_{r_{\rm s}} &\simeq& \frac{M_{\bullet}(2-\Gamma)}{4\pi r_{\rm soi}^{3}}\left(\frac{r_{\rm s}}{r_{\rm soi}}\right)^{-1-\Gamma} \nonumber \\
 &\approx& \begin{cases}
    9.4 \times 10^{-19}M_{\bullet,8}^{-1.2}v_{500}^{2.2}\,{\rm g\,cm^{-3}}
    & \text{core} \\
    1.1 \times 10^{-18} M_{\bullet,8}^{-1.5}v_{500}^{3.6}
    \,{\rm g\,cm^{-3}}  & \text{cusp}, 
  \end{cases}
  \label{eq:rhostarrs}
\end{eqnarray}
Even in the saturated case, heat conduction is at most an order unity correction for expected values of the saturation parameter $\phi < 0.1$, justifying its neglect in our analysis.

@@

If important, radiative cooling qualitative alters key features of the accretion flow.  Radiative cooling is usually of greatest important near the stagnation radius, where the gas is in nearly hydrostatic balance.  Initially hydrostatic gas that cools at constant pressure $P \propto nT$ at the rate $\dot{q}_{\rm rad} = n^{2}\Lambda \propto T^{-2.7}$ (eq.~[\ref{eq:qdotrad}]) is thermally unstable, where $\Lambda(T) \propto T^{-\beta}$ with $\beta \sim -0.7$ is the cooling function (see below).  Thermal instability leads to initially higher black hole accretion rates and enhanced star formation which, when coupled to the flow by feed-back processes, can lead to limit cycle behavior inconsistent with our assumption of steady flow (e.g.~\citealt{Ciotti&Ostriker07}; \citealt{Ciotti+10}; \citealt{Gan+14}).

>>>>

Radiative cooling contributes an additional term to equation (\ref{eq:dsdt}) of the form
\be
\dot{q}_{\rm rad} = -\Lambda(T)n^{2},
\label{eq:qdot_rad}
\ee
where $n = \rho/\mu m_p$ is the gas number density and $\Lambda(T)$ is the cooling function.  We neglect radiative cooling in our baseline calculations, despite the fact that this is not justified when the wind heating $\vwO$ is low or if the mass return rate $\eta$ is high.  However, once radiative cooling becomes important compared to other sources of heating and cooling, its presence leads to thermal instability (e.g.~\citealt{McCourt+12}) that cannot be accounted for by our 1D
time-independent model.  A main goal of this work is to use solutions neglecting radiation to ascertain over what range of physical conditions cooling instabilities are most likely to develop ($\S\ref{sec:cooling}$).

@@

In the opposite limit of weak heating ($v_{w} \ll \sigma$), the stagnation radius moves to large radii $\gg r_{\rm soi}$ or may not exist at all, formally implying that the entire ISM of the galaxy is accreting.  This result makes intuitive sense: gas is supplied to the nuclear by stars which are gravitationally bound to the black hole, so outflows are possible only if the specific heating rate $\sim v_{w}^{2}$ significantly exceeds the specific gravitational binding energy $\sim \sigma^{2}$.

@@

The accretion rate onto the SMBH is given by the integrated mass loss rate interior to the stagnation radius (eq.~[\ref{eq:q}]), 
\begin{eqnarray}
  \dot{M} &=& \frac{\eta \Menc|_{r_{\rm s}}}{\th} \nonumber \\
&\approx&
  \begin{cases}
    2.9 \times 10^{-5} M_{\bullet,8}^{1.76}
    v_{500}^{-3.8}  \eta_{0.02} \Msun \, \pyear& \text{core} \\
    5.3 \times 10^{-5} M_{\bullet,8}^{1.48} 
    v_{500}^{-2.4}  \eta_{0.02} \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}
where we have assumed $v_{w} \gg \sigma$ by adopting equation (\ref{eq:stag_simple}) for $r_s$, and we have divided our results into core ($\Gamma = 0.1$) and cusp ($\Gamma = 0.8$) galaxies.  The resulting Eddington ratio is given by 
\begin{eqnarray}
\frac{\dot{M}}{\dot{M}_{\rm edd}} &\approx&
  \begin{cases}
    1.3 \times 10^{-5} M_{\bullet,8}^{0.76}
    v_{500}^{-3.8}  \eta_{0.02}   & \text{core}, \\
    2.4 \times 10^{-5} \Mbheight^{0.48} 
    v_{500}^{-2.4}  \eta_{0.02}   & \text{cusp}, 
  \end{cases}
  \label{eq:eddr_analytic}
\end{eqnarray}
where $\dot{M}_{\rm edd} = 2.6M_{\bullet,8}M_{\odot}$ yr$^{-1}$  is the Eddington accretion rate, assuming a radiative efficiency of ten per cent.  Note the sensitive inverse dependence of the accretion rate on the wind heating rate.  
%To reproduce the estimated values $\dot{M}/\dot{M}_{\rm edd} \lesssim 10^{-3}$ in low-luminosity AGN or quiescent galaxies, we thus require $v_{w} \gtrsim 150M_{\bullet,8}^{0.5}$

@@

Our expression for the mass accretion rate (eq.~[\ref{eq:mdot_analytic}]) can be compared to the traditional Bondi rate for accretion onto a point source from an external medium of specified density and temperature (\citealt{Bondi52}):
\begin{align}
  \dot{M}_{\rm B} =4\pi \lambda r_{\rm B}^2 \rho|_{r_{\rm B}}v_{\rm ff}|_{r_{\rm B}},
\label{eq:bondi}
\end{align}
where $r_{\rm B} \equiv GM/c_{\rm s,ad}^{2}$ is the Bondi radius, $c_{\rm s,ad} = \sqrt{\gamma kT/\mu m_p}$ is the adiabatic sound speed, $v_{\rm ff}|_{r_{\rm B}} = r_{\rm B}/t_{\rm ff}|_{r_{\rm B}} = (GM_{\bullet}/r_{\rm B})^{1/2}$ and $\lambda$ is a parameter of order unity.  One observes a similarity between equation (\ref{eq:mdot_gas}) and the Bondi formula (eq.~[\ref{eq:bondi}]) if $r_{\rm B}$ is replaced by $\rs$.  Indeed, when the stagnation radius resides interior to the sphere of influence we find that (eq.~[\ref{eq:stag_simple}])
\begin{align}
  \rs\approx\frac{7}{2}\frac{G \Mbh}{n \vwO^2} \approx \frac{7}{6n}r_{\rm B},
  \label{eq:rs_simple}
\end{align}
where the second equality makes use of equation (\ref{eq:Tanalytic}).  {\bf BDM: please check, as I found something different than what was here before}.

@@

Our analytic derviations neglect the effects of heat conduction, an assumption we now check.  The ratio of the magnitude of the conductive heating rate to the external heating rate at the stagnation radius is given by
\begin{align}
  \left.\frac{\nabla\cdot(\kappa \nabla T)}{q v_{\rm
w}^{2}/2}\right|_{r_{\rm s}} &\sim \frac{2t_{\rm h}\kappa_0
T|_{r_{\rm s}}^{7/2}}{r_{\rm s}^{2}\eta \rho_{\star}|_{r_{\rm s}} \tilde{v}_{w}^{2}(1+\phi)}
\nonumber \\ &\sim {\rm min}
  \begin{cases}
  10 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.8} v_{500}^{6.8} &  \text{unsaturated (core)}\\
 10 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.5} v_{500}^{5.4} &  \text{unsaturated (cusp)}\\
  1.0(\phi/0.1)/(2-\Gamma) & \text{saturated},
  \end{cases}
 \label{eq:conduction}
\end{align}
where the second equality makes use of equations (\ref{eq:Tanalytic}), (\ref{eq:stag_simple}), and we have approximated $\nabla^{2} \sim 1/r_{\rm s}^{2}$, $\nabla \sim 1/r_{\rm s}$.  We have approximated the stellar density profile as (eq.~[\ref{eq:rhostar}])
\begin{eqnarray}
  \rho_{\star}|_{r_{\rm s}} &\simeq& \frac{M_{\bullet}(2-\Gamma)}{4\pi r_{\rm soi}^{3}}\left(\frac{r_{\rm s}}{r_{\rm soi}}\right)^{-1-\Gamma} \nonumber \\
 &\approx& \begin{cases}
    9.4 \times 10^{-19}M_{\bullet,8}^{-1.2}v_{500}^{2.2}\,{\rm g\,cm^{-3}}
    & \text{core} \\
    1.1 \times 10^{-18} M_{\bullet,8}^{-1.5}v_{500}^{3.6}
    \,{\rm g\,cm^{-3}}  & \text{cusp}, 
  \end{cases}
  \label{eq:rhostarrs}
\end{eqnarray}
where we have assumed that the stagnation radius is well inside the Nuker break radius.  Equation (\ref{eq:conduction}) shows that even in the saturated case, neglecting heat conduction results in at most an order unity error in the entropy source term for expected values of the saturation parameter $\phi < 0.1$, justifying its neglect in our analysis.  Numerical experiments including conductivity are seen confirm this ($\S\ref{sec:numerical}$).

@@

If radiative cooling becomes important, it can qualitatively alter key features of the accretion flow.  Radiative cooling usually has its greatest impact near or external to the stagnation radius, where the gas resides in near hydrostatic balance.  Initially hydrostatic gas that cools at constant pressure $P \propto nT$ at the rate $\dot{q}_{\rm rad} = n^{2}\Lambda \propto T^{-2.7}$ (eq.~[\ref{eq:qdotrad}]) is thermally unstable, where $\Lambda(T) \propto T^{-\beta}$ with $\beta \sim -0.7$ is the cooling function (see below).  Thermal instability leads to initially higher black hole accretion rates and enhanced star formation which, when coupled to the flow by feed-back processes, can lead to limit cycle behavior inconsistent with our assumption of steady flow (e.g.~\citealt{Ciotti&Ostriker07}; \citealt{Ciotti+10}; \citealt{Gan+14}).
